Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2022-02-10 paper/paper.tex
WIP moving to laptop

\color{grey}
\textbf{New introduction:} Simpson and Goodhill presented a model in which the
chemoaffinity mechanism is non-biological. We offer a model in which a
competition mechanism counteracts the drive of axons along the ephrin
gradient. Will we need the axon-axon interaction mechanism based on
non-same-Eph repulsion?

\textbf{Another approach:} We present a model of gradient-following based on graded
interactions between the growth cones of retinal ganglion cell axons and the
tectal surface. We show that a combination of exponentially graded expression
in the retina \citep{reber_relative_2004} and exponentially (for repulsively
interacting receptor ligand pairs) \emph{or} logarithmically (for attractively
interacting receptor/ligands) graded expression on the tectum permits a purely
local model to reproduce the majority of reported phenomena associated with
retinotectal organisation.

There have been many approaches to modelling the organisation of the
retinotectal projection. Many of these models are wholly or partially
phenomenological.

@@

\color{grey}
This means that growth cones of approximate
radius 5 $\mu$m = 0.005 mm \citep{goodhill_can_2016} have a radius $r$ =
0.0025 in the arbitrary units of the simulation. For a realistic simulation
based on mouse, the approximate RGC density in the retina is around 44000
cells per retina \citep{jeon_major_1998}. This sets the side length of the
retina to $\sqrt{44000} \approx 200$. This results in a simulation of
prodigious computational size, which although tractable (10 s per step,
requiring several hours, rather than days to run) is not convenient. To
maintain the relationship between growth cone size and cell density, we permit
a smaller number of growth cones ($400\times8=3200$) but keep the fractional
area of the branches at 0.86 ($44000 \times \pi r^2$) (if there is one cone
per axon) or 6.9 (if there are 8 cones per axon).

\color{black}
Each discrete, square element in Fig.\,\ref{f:ex}B projects one RGC axon with $M=4$
branches to the tectum, carrying the receptor expression at the retinal
element's location.
At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, at time
$t$ of each branch $b$ is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the branch movement vector, is the sum of a chemoaffinity
effect and a `border effect', $\mathbf{B}_b$, which acts to retain branches
within the tectal region:
%
\begin{equation} \label{e:mv}
\mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + 0.5 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$.

@@

\citet{simpson_simple_2011} showed that repulsive interactions combined with a
chemoaffinity-like effect could account for the surgical manipulations shown
in Figs.\,\ref{f:trot90}--\ref{f:Gsurg}. To confirm that a competitive
mechanism could operate successfully in combination with the
gradient-following chemoaffinity mechanism described above, we introduced
three competition mechanisms and investigated their behaviour.

These were: space-based competition (`$\mathbf{C}$'), as described
in \citet{simpson_simple_2011}; a receptor-receptor signalling mechanism
(`$\mathbf{I}$'), similar to that described in \citet{simpson_simple_2011},
and a receptor-ligand mass-action signalling model
(`$\mathbf{J}$'). The movement vector (Eq.\,\ref{e:mv}) becomes
%
\begin{equation} \label{e:mv2}
\mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b +  m_c \mathbf{C}_b +  m_{\!_I} \mathbf{I}_b +  m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
\end{equation}
%

Although two of these competitive mechanisms were investigated
by \citet{simpson_simple_2011}, the first (space-based competition) offers no
mechanism by which axon growth cones are able to communicate and the second
mechanism was implemented only for a single axis in the model (EphA receptors,
expressed in a gradient along the XX retinal axis). We wondered whether a
\emph{single} signalling mechanism working alongside the `untuned'
chemoaffinity with ligand expression given by Eq.\,\ref{e:tecligexp} could
successfully reproduce the wildtype and manipulation results. Alongside the
same chemoaffinity mechanism, we studied each competitive mechanism in turn,
setting $m_{\!_I}=m_{\!_J}=0$ to give a `$\mathbf{GC}$' model,
$m_c=m_{\!_J}=0$ to give a `$\mathbf{GI}$' model or $m_c=m_{\!_I}=0$ to give a
`$\mathbf{GJ}$' model.

\subsection*{Chemoaffinity plus space-based competition (GC)}

Space based competition was implemented as a repulsive effect acting between
any two branches, whether they originated from different retinal ganglion
cells or from the same cell. An assumption is that there is some kind of
self-other signal which acts to discourage two cells from approaching each
other. A competition movement vector, $\mathbf{C}_b$, acting on branch $b$, is
defined as a proximity-weighted sum of the repulsion due to all nearby branches:
%
\begin{equation} \label{e:C}
\mathbf{C}_b = \frac{1}{|B_{b,c}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb} W_c
\end{equation}
where $\hat{\mathbf{x}}_{kb}$ is the unit vector from branch $k$ to branch
$b$. The distance-dependent weight $W_c$ is given by
\begin{equation}
W_c(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_c}   & d_{kb} \leq 2r_c \\
     0 & d_{kb} > 2r_c
     \end{cases}
\end{equation}
for two growth cones of interaction radius $r_c$ separated by a distance
$d_{kb}$. $B_{b,c}$ is the set of growth cones within a distance $2r_c$ of
branch $b$ and the total number of branches in the system is the number of RGC
axons, $n$ multiplied by the number of branches per axon, $M$. This
description of the competition matches that in \citet{simpson_simple_2011}.

@@

The model defined by the receptor and ligand expressions of Eqs.\,\ref{e:retrcpt}
\& \ref{e:tecligexp} along with Eq.\,\ref{e:mv2} with $m_{\!_I}=m_{\!_J}=0$ has three
free parameters; $m_{\!_G}$, $m_c$ and $r_c$. To find the best combination of
these parameters, we reasoned that evolution would optimize the arrangement of
the wildtype pattern. We performed a simulated annealing optimization to
minimize the error metric, $\epsilon$, for the wildtype pattern
(Fig.\,\ref{f:G}E) at simulation time 1000. The parameters returned were
$m_{\!_G}$ = 0.003841, $m_c$ = 0.09959 and $r_c$ = 0.39918. Fig.\,\ref{f:GC}
shows the $\mathbf{GC}$ simulation with these parameters. Fish net plots at
$t=40$ and $t=1000$ show that the pattern develops quickly and compares well
with the expected layout (Fig.\,\ref{f:G}E). We investigated the response of
this model to the same set of surgical manipulations shown for the
chemoaffinity-only model in Fig.\,\ref{f:Gsurg}. We found that the addition of
competition leads to a better correspondence to the experimental results than
for the chemoaffinity-only model (see supplementary Fig.\,\ref{f:GCsurg}).

\subsection*{Receptor-ligand axon-axon interactions (GJ)}

Because the competition effect, $\mathbf{C}_b$ (Eq.\,\ref{e:C}), requires a
global supervisor to provide information to each branch about the location of
all other branches, we looked for a receptor-ligand interaction that could
signal between axon branches and enable them to adjust their movements away
from one another. We investigated the idea that the same receptor-ligand
interaction that is modelled between RGC axons and the tectum could also act
as a signal between the axons, because retinal ganglion cells are known to
express ephrin receptor ligands as well as Eph
receptors \citep{ephrins_on_RGCs}. We defined $N=4$ ligands, $l_i$, expressed
by the RGCs in the model according to
%
\begin{equation} \label{e:retlig}
l_i(x,y) = 1.05 + 0.26 \exp(2.3 u).
\end{equation}
%
We used exactly the same form as the receptor expression given by
Eq.\,\ref{e:retrcpt}, but with opposite gradient direction so here $u$ is
substituted by $x$ for $l_0$, $y$ for $l_1$, $(1-x)$ for $l_2$ and by $(1-y)$
for $l_3$ (see Fig.\,\ref{f:ex}C). We modelled a
multiplicative (or mass-action?) interaction to give the movement
vector, $\mathbf{J}_b$, as  follows:
%
\begin{equation}
% No \frac{1}{|B_b|} in this one. Really?
\mathbf{J}_b = \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,Q_{\!_J}(d_{kb})
\end{equation}
%
where $Q_{\!_J}(d_{kb})$ is the signalling strength between two growth cones of
receptor-ligand interaction radius $r_{\!_J}$ a distance $d_{kb}$ from one another,
given by
%
\begin{equation}
Q_{\!_J}(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r_{\!_J} \\
     0 & d_{kb} > 2r_{\!_J}
     \end{cases}
\end{equation}
%
The sign of $Q_{\!_J}$ (which is dependent on the values of $F_i$) determines
whether the interaction, $\mathbf{J}_b$, is repulsive or attractive. $l_{i,k}$
is the expression of ligand $i$ on branch $k$.

@@

We performed the same simulated annealing process as for the
$\mathbf{GC}$-model to find the parameters best able to recover the wildtype
pattern and found $m_{\!_G} = 0.00047$, $m_{\!_J} = 0.00089$ and $r_{\!_J} = 0.341$.

@@

\subsection*{Receptor-receptor axon-axon interactions (GI)}

>>>>

%% \color{grey}
%% \textbf{New introduction:} Simpson and Goodhill presented a model in which the
%% chemoaffinity mechanism is non-biological. We offer a model in which a
%% competition mechanism counteracts the drive of axons along the ephrin
%% gradient. Will we need the axon-axon interaction mechanism based on
%% non-same-Eph repulsion?

%% \textbf{Another approach:} We present a model of gradient-following based on graded
%% interactions between the growth cones of retinal ganglion cell axons and the
%% tectal surface. We show that a combination of exponentially graded expression
%% in the retina \citep{reber_relative_2004} and exponentially (for repulsively
%% interacting receptor ligand pairs) \emph{or} logarithmically (for attractively
%% interacting receptor/ligands) graded expression on the tectum permits a purely
%% local model to reproduce the majority of reported phenomena associated with
%% retinotectal organisation.

%% There have been many approaches to modelling the organisation of the
%% retinotectal projection. Many of these models are wholly or partially
%% phenomenological.

@@

%% This means that growth cones of approximate
%% radius 5 $\mu$m = 0.005 mm \citep{goodhill_can_2016} have a radius $r$ =
%% 0.0025 in the arbitrary units of the simulation. For a realistic simulation
%% based on mouse, the approximate RGC density in the retina is around 44000
%% cells per retina \citep{jeon_major_1998}. This sets the side length of the
%% retina to $\sqrt{44000} \approx 200$. This results in a simulation of
%% prodigious computational size, which although tractable (10 s per step,
%% requiring several hours, rather than days to run) is not convenient. To
%% maintain the relationship between growth cone size and cell density, we permit
%% a smaller number of growth cones ($400\times8=3200$) but keep the fractional
%% area of the branches at 0.86 ($44000 \times \pi r^2$) (if there is one cone
%% per axon) or 6.9 (if there are 8 cones per axon).

Each discrete, square element in Fig.\,\ref{f:ex}B projects one RGC axon with $M=4$
branches to the tectum, carrying the receptor expression at the retinal
element's location.
At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, at time
$t$ of each branch $b$ is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the branch movement vector, is the sum of a chemoaffinity
effect and a `border effect', $\mathbf{B}_b$, which acts to retain branches
within the tectal region:
%
\begin{equation} \label{e:mv}
\mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + 0.5 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$.

@@

\citet{simpson_simple_2011} showed that repulsive interactions combined with a
chemoaffinity-like effect could account for the surgical manipulations shown
in Figs.\,\ref{f:trot90}--\ref{f:Gsurg}. To confirm that a competitive
mechanism could operate successfully in combination with the
gradient-following chemoaffinity mechanism described above, we first
implemented the `space based' competition of \citet{simpson_simple_2011},
updating the movement vector with a term $m_c \mathbf{C}_b$:
\begin{equation} \label{e:mv2}
 \mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b +  m_c \mathbf{C}_b + 0.5 \mathbf{B}_b
 %%+  m_{\!_I} \mathbf{I}_b +  m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
\end{equation}
where $m_c$ is simply a scalar parameter.
%
%
%% introduced
%% three competition mechanisms and investigated their behaviour.
%
%% These were: space-based competition (`$\mathbf{C}$'), as described
%% in \citet{simpson_simple_2011}; a receptor-receptor signalling mechanism
%% (`$\mathbf{I}$'), similar to that described in \citet{simpson_simple_2011},
%% and a receptor-ligand mass-action signalling model
%% (`$\mathbf{J}$'). The movement vector (Eq.\,\ref{e:mv}) becomes
%% %
%% \begin{equation} \label{e:mv2}
%% \mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b +  m_c \mathbf{C}_b +  m_{\!_I} \mathbf{I}_b +  m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
%% \end{equation}
%% %
%
%% Although two of these competitive mechanisms were investigated
%% by \citet{simpson_simple_2011}, the first (space-based competition) offers no
%% mechanism by which axon growth cones are able to communicate and the second
%% mechanism was implemented only for a single axis in the model (EphA receptors,
%% expressed in a gradient along the XX retinal axis). We wondered whether a
%% \emph{single} signalling mechanism working alongside the `untuned'
%% chemoaffinity with ligand expression given by Eq.\,\ref{e:tecligexp} could
%% successfully reproduce the wildtype and manipulation results. Alongside the
%% same chemoaffinity mechanism, we studied each competitive mechanism in turn,
%% setting $m_{\!_I}=m_{\!_J}=0$ to give a `$\mathbf{GC}$' model,
%% $m_c=m_{\!_J}=0$ to give a `$\mathbf{GI}$' model or $m_c=m_{\!_I}=0$ to give a
%% `$\mathbf{GJ}$' model.
%
%%\subsection*{Chemoaffinity plus space-based competition (GC)}
%
The competition movement vector, $\mathbf{C}_b$, introduces a repulsive effect
acting between any two branches, whether they originated from different
retinal ganglion cells or from the same cell. An assumption is that there is
some kind of self-other signal which acts to discourage two cells from
approaching each other. $\mathbf{C}_b$ is defined as a proximity-weighted sum
of the repulsion due to all nearby branches:
%
\begin{equation} \label{e:C}
\mathbf{C}_b = \frac{1}{|B_{b,c}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb} W_c
\end{equation}
where $\hat{\mathbf{x}}_{kb}$ is the unit vector from branch $k$ to branch
$b$. The distance-dependent weight $W_c$ is given by
\begin{equation}
W_c(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_c}   & d_{kb} \leq 2r_c \\
     0 & d_{kb} > 2r_c
     \end{cases}
\end{equation}
for two growth cones of interaction radius $r_c$ separated by a distance
$d_{kb}$. $B_{b,c}$ is the set of growth cones within a distance $2r_c$ of
branch $b$ and the total number of branches in the system is the number of RGC
axons, $n$ multiplied by the number of branches per axon, $M$. This
description of the competition matches that
in \citet{simpson_simple_2011}. Because the model incorporates a chemoaffinity
effect and a competition effect, we called it the `$\mathbf{GC}$ model'.

@@

The model defined by the receptor and ligand expressions of Eqs.\,\ref{e:retrcpt}
\& \ref{e:tecligexp} along with Eq.\,\ref{e:mv2} has three
free parameters; $m_{\!_G}$, $m_c$ and $r_c$. To find the best combination of
these parameters, we reasoned that evolution would optimize the arrangement of
the wildtype pattern. We performed a simulated annealing optimization to
minimize the error metric, $\epsilon$, for the wildtype pattern
(Fig.\,\ref{f:G}E) at simulation time 1000. The parameters returned were
$m_{\!_G}$ = 0.003841, $m_c$ = 0.09959 and $r_c$ = 0.39918. Fig.\,\ref{f:GC}
shows the $\mathbf{GC}$ simulation with these parameters. Fish net plots at
$t=40$ and $t=1000$ show that the pattern develops quickly and compares well
with the expected layout (Fig.\,\ref{f:G}E). We investigated the response of
this model to the same set of surgical manipulations shown for the
chemoaffinity-only model in Fig.\,\ref{f:Gsurg}. We found that the addition of
competition leads to a better correspondence to the experimental results than
for the chemoaffinity-only model (see supplementary Fig.\,\ref{f:GCsurg}).

\subsection*{Receptor-ligand axon-axon interactions}

Because the competition effect, $\mathbf{C}_b$ (Eq.\,\ref{e:C}), requires a
global supervisor to provide information to each branch about the location of
all other branches, we looked for a receptor-ligand interaction that could
signal between axon branches and enable them to adjust their movements away
from one another. We investigated the idea that the same receptor-ligand
interaction that leads to gradient-based movement of RGC axons on the tectum
could also act as a signal between the axons. This is plausible, because
retinal ganglion cells are known to express ephrin receptor ligands as well as
Eph receptors \citep{ephrins_on_RGCs}. We defined $N=4$ ligands, $l_i$,
expressed by the RGCs in the model according to
%
\begin{equation} \label{e:retlig}
l_i(x,y) = 1.05 + 0.26 \exp(2.3 u).
\end{equation}
%
We used exactly the same form as the receptor expression given by
Eq.\,\ref{e:retrcpt}, but with opposite gradient direction so here $u$ is
substituted by $x$ for $l_0$, $y$ for $l_1$, $(1-x)$ for $l_2$ and by $(1-y)$
for $l_3$ (see Fig.\,\ref{f:ex}C).
%
We combined chemotaxis with a multiplicative (or mass-action?) interaction,
$\mathbf{J}_b$ (scaled by a parameter $m_{\!_J}$) to define the movement
vector, $\mathbf{M}_b$, for the $\mathbf{GJ}$ model as:
%
\begin{equation} \label{e:mv3}
 \mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
 %%+  m_{\!_I} \mathbf{I}_b +  m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
\end{equation}
%
$\mathbf{J}_b$ is defined as
%
\begin{equation}
% No \frac{1}{|B_b|} in this one. Really?
\mathbf{J}_b = \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,Q_{\!_J}(d_{kb})
\end{equation}
%
where $Q_{\!_J}(d_{kb})$ is the signalling strength between two growth cones of
receptor-ligand interaction radius $r_{\!_J}$ a distance $d_{kb}$ from one another,
given by
%
\begin{equation}
Q_{\!_J}(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r_{\!_J} \\
     0 & d_{kb} > 2r_{\!_J}
     \end{cases}
\end{equation}
%
The sign of $Q_{\!_J}$ (which is dependent on the values of $F_i$) determines
whether the interaction, $\mathbf{J}_b$, is repulsive or attractive. $l_{i,k}$
is the expression of ligand $i$ on branch $k$.

@@

We performed the same simulated annealing process as for the
$\mathbf{GC}$ model to find the parameters best able to recover the wildtype
pattern and found $m_{\!_G} = 0.00047$, $m_{\!_J} = 0.00089$ and $r_{\!_J} =
0.341$.

@@

\subsection*{Receptor-receptor axon-axon interactions}
